---
title: "Penguin Classification"
author: "Joan Balaguer, Jaume Adrover"
output: html_document
date: '2023-12-08'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(caret)
library(cluster)    
library(factoextra)
library(dendextend)
library(stats)
```

## 0. Introduction

In this assignment, our goal is to find out as many possible ways to group  penguin according to different features, such as bill length or mass weight.

We'll accomplish such task by using different clustering methods. These clustering methods will be grouped in hierarchical and partitioning ones.


## 1. Exploratory data analysis

First of all, let's load our data and show a little summary, along with the head function to display few rows.

```{r}
data <- read.csv("penguindata.csv")
head(data)
```

```{r}
summary(data)
```

### 1.2 Plotting single data values
In this stage, we're going to see how variables are built in our dataset so we can group in a better way.

#### 1.2.1 Bill depth
This numerical column expresses in millimeters how deep a penguin's bill is. Let's see its distribution:

```{r}
hist(data$bill_depth_mm,main="Bill depth distribution",xlab="Bill depth(mm)",col="cyan")
```

#### 1.2.2 Bill length
This numerical column expresses in millimeters how long a penguin's bill is. Let's see its distribution:

```{r}
hist(data$bill_length_mm,main="Bill length distribution",xlab="Bill length(mm)",col="red")
```

It seems bill length is equally distributed in the 37-55 mm range. 

#### 1.2.3 Flipper length

Next, we're going to look forward to how flipper length is distributed.

```{r}
hist(data$flipper_length_mm,main="Flipper length distribution",xlab="Flipper length(mm)",col="brown")
```

It seems most penguins have 185-195 mm flippers. 

#### 1.2.4 Body mass

```{r}
hist(data$body_mass_g,main="Body mass distribution",xlab="Body mass(g)",col="brown")

ggplot(data, aes(x = body_mass_g, fill = "brown")) +
  geom_density(alpha = 0.7) +  # Add alpha for transparency
  labs(title = "Density Plot of Body Mass", x = "Body Mass (g)") +
  theme_minimal() +
  theme(legend.position = "none")
```


#### 1.2.5 Sex
  Considering sex is a categorical column, we'll show it with a pie chart and also a histogram. Let's see % of each gender group

```{r}
# Gender Pie Chart
ggplot(data, aes(x = "", fill = sex)) +
  geom_bar(width = 1, position = "fill") +
  coord_polar(theta = "y") +
  labs(title = "Gender Pie Chart") +
  theme_minimal()
```

As we can clearly see, it seems that gender distribution is equal, without considering NA values that will be removed later. Let's compare with a different point of view:

```{r}
# Gender Histogram
ggplot(data, aes(x = sex, fill = sex)) +
  geom_bar() +
  labs(title = "Gender histogram") +
  theme_minimal()
```

This different plot allows us to see that male count is slightly higher.

#### 1.2.6 Study year

```{r}
# Year histogram
ggplot(data, aes(x = year, fill = year)) +
  geom_bar() +
  labs(title = "Year histogram") +
  theme_minimal()
```

## 2. Data cleaning
In this data cleaning phase, we'll basically take care of these cases:

1. Missing values
2. Filter relevant data
3. Categorical columns
4. Data Scaling


### 2.1 Missing values
Let's look for missing values with this function:
```{r cars}
na_summary_func <- function(data) {
  # NA values for column
  na_counts <- colSums(is.na(data))

  na_summary <- data.frame(Column = names(na_counts), NA_Count = na_counts)
  
  na_summary <- na_summary[na_summary$NA_Count > 0,]

  return(na_summary)
}
```

Let's use it.
```{r}
na_summary_func(data)
```
As we can clearly see, there are 11 rows where sex is NA and other numeric values in 2 rows. Considering we can't replace sex, we are going to remove all rows where there is any NA value.

If we only had missing values in numerical columns one choice would be to replace them for the mean.

```{r}
clean_data <- na.omit(data)
```

Let's check if it worked.

```{r}
na_summary_func(clean_data)
```
We have 0 rows with missing values, great.

### 2.2 Filter relevant data
As we have seen, the only non desired column would be X, as it acts as an unique identifier for each row.


```{r}
clean_data <- subset(clean_data, select = -X)
```




### 2.3 Categorical columns
We could consider year a categorical column but it is formatted as a numerical one son we'll only focus on sex one.

We are going to replcae unique values like this:

1- Male:   1
2- Female: 0

```{r}
clean_data$sex <- ifelse(clean_data$sex=="male",1,0)

```

Let's check:

```{r}
head(clean_data)
```
### 2.4 Data scaling
In this final data preparation step we're going to normalize values between 0 and 1.

We'll use caret built-in function preProcess that uses this formula:
  
\[ x' = \frac{{x - \text{min}(X)}}{{\text{max}(X) - \text{min}(X)}} \]

Result from this formula meets the following condition:

\[ x' \in [0,1] \]

Let's scale our values:

```{r}
preprocess_model <- preProcess(clean_data, method = "range")

# Scale whole dataframe
df_scaled <- predict(preprocess_model, newdata = clean_data)

summary(df_scaled)
```
## 3. Clustering


### 3.1 Plotting relationship between variables

Now, we're going to consider different variables pair so we can choose which combinations fit the best to group our beloved penguins.
We'll plot different ones and choose what we think are the best pairs. We have to exclude year as we think it's not a great grouping variable as it's not implicit en penguin features. Our goal would be to group new penguins into clusters and study year couldn't be included.

#### 3.1.1 Sex

Sex vs body mass:
```{r}
plot(df_scaled$sex, df_scaled$body_mass_g, 
     xlab = "Sex", 
     ylab = "Body Mass",
     main = "Sex vs. Body Mass")
```

We can clearly see that we could pull apart 4 different groups and that male gender is associated with a higher body mass.

Sex vs Flipper length:
```{r}
plot(df_scaled$sex, df_scaled$flipper_length_mm, 
     xlab = "Sex", 
     ylab = "Flipper length",
     main = "Sex vs Flipper length")
```

This could be a possible grouping criteria, it seems male penguins have longer flippers.

Sex vs bill length:
```{r}
plot(df_scaled$sex, df_scaled$bill_length_mm, 
     xlab = "Sex", 
     ylab = "Bill length",
     main = "Sex vs Bill length")
```


Sex vs bill depth:

```{r}
plot(df_scaled$sex, df_scaled$bill_depth_mm, 
     xlab = "Sex", 
     ylab = "Bill depth",
     main = "Sex vs Bill depth")
```

#### 3.1.2 Bill length

Bill length vs Body Mass:

```{r}
plot(df_scaled$bill_length_mm, df_scaled$body_mass_g, 
     xlab = "Bill length", 
     ylab = "Body Mass",
     main = "Bill length vs Body Mass")
```



Bill length vs Flipper length:

```{r}
plot(df_scaled$bill_length_mm, df_scaled$flipper_length_mm, 
     xlab = "Bill length", 
     ylab = "Flipper Length",
     main = "Bill length vs Flipper Length")
```



Bill length vs Bill depth:

```{r}
plot(df_scaled$bill_length_mm, df_scaled$bill_depth_mm, 
     xlab = "Bill length", 
     ylab = "Bill depth",
     main = "Bill length vs Bill Depth")
```

#### 3.1.3 Bill depth

Bill depth vs Body Mass:

```{r}
plot(df_scaled$bill_depth_mm, df_scaled$body_mass_g, 
     xlab = "Bill Depth", 
     ylab = "Body Mass",
     main = "Bill Depth vs Body Mass")
```

Bill depth vs Flipper Length:

```{r}
plot(df_scaled$bill_depth_mm, df_scaled$flipper_length_mm, 
     xlab = "Bill Depth", 
     ylab = "Flipper Length",
     main = "Bill Depth vs Flipper Length")
```

#### 3.1.4 Flipper length

Flipper length vs Body Mass

```{r}
plot(df_scaled$flipper_length_mm, df_scaled$body_mass_g, 
     xlab = "Flipper Length", 
     ylab = "Body Mass",
     main = "Flipper Length vs Body Mass")
```

#### 3.1.5 Chosen combinations

Final list is:

According to *sex*:

1. Sex vs Body Mass
2. Sex vs Bill Depth
3. Sex vs Bill Length
4. Sex vs Flipper Length

According to *flipper length*:


1. Flipper Length vs Bill Length
2. Flipper Length vs Bill Depth
3. Flipper Length vs Body Mass

According to *bill length*:

1. Bill Length vs Body Mass
2. Bill Length vs Bill Depth


These are going to be some of our final 2x variable combinations to cluster by.

### 3.2 Partitioning Clustering models

#### 3.2.1 K-means

First of all, we are going to make an agroupment, creating a clustering k-means model for each combination.
Let's define all necessary functions:

```{r}
# Function to create Cluster given k,nstart and specified columns
createCluster <- function(k_value,nstart_value,cols){
  set.seed(123)
  df = df_scaled[cols]

  # K-means
  kmeans_result <- kmeans(df, centers = k_value, nstart = nstart_value)
  
  # Cluster plotting
  kmeans_plot <-fviz_cluster(kmeans_result, data = df)
  
  return (list(result=kmeans_result,plot=kmeans_plot))
}

# Function to plot cluster means
cluster_means <-function(kmeans,cols){
  for (i in 1:nrow(kmeans)){
    cat("Cluster", i - 1, "->",cols[1],":", kmeans[i, cols[1]], ",", cols[2],":", kmeans[i, cols[2]], "\n")
  }
}
```


##### **Sex vs body_mass_g**

Let's try with a k=3 model:
```{r}
cols=c("sex","body_mass_g")
cluster <- createCluster(3,25,cols)
cluster$plot
```

It kinda looks like there could be two groups in the left column, let's increase k by 1.
```{r}
cluster <- createCluster(4,25,cols)
cluster$plot
```

That looks much better, let's see every cluster mean:
```{r}
# Show cluster means
cluster_means(cluster$result$centers,cols)
```

##### **Sex vs bill_depth_mm**

Let's change our column choice now.
```{r}
cols=c("sex","bill_depth_mm")
cluster <- createCluster(3,25,cols)
cluster$plot
```

It seems that clusters are a little big, they keep lot of points, let's try a bigger k value.

```{r}
cluster <- createCluster(4,25,cols)
cluster$plot
```

That looks much better, these could behave as 4 different groups:

1. Small depth bill,females(cluster 2)
2. Large depth bill,females(cluster 1)
3. Small depth bill,males(cluster 3)
4. Large depth bill,males(cluster 4)

And, given these grouping criteria, we could almost add a medium category to each gender, so we would have k=6:
```{r}
cluster <- createCluster(6,25,cols)
cluster$plot
```

Then, we would have previous categories plus:

5. Medium depth bill, females
6. Medium depth bill, males


Let's finally see every cluster mean:
```{r}
cluster_means(cluster$result$centers,cols)
```

That could be a nice grouping rule.


##### **Sex vs bill_length_mm**

Let's apply the same grouping rule as the previous one. Let's start with k=4.
```{r}
cols=c("sex","bill_length_mm")
cluster <- createCluster(4,25,cols)
cluster$plot
```

We could say the same about this one, let's add a medium category, so k=6
```{r}
cluster <- createCluster(6,25,cols)
cluster$plot
```

We now have 3 categories for each gender: small,medium and large long bills.


Let's see every cluster mean:
```{r}
cluster_means(cluster$result$centers,cols)
```

That would also be a nice,similar grouping rule.

##### **Sex vs flipper_length_mm**
Considering flipper length is similar to previous columns, we'll initially try with k=6.

```{r}
cols=c("sex","flipper_length_mm")
cluster <- createCluster(6,25,cols)
cluster$plot
```

This is great, it works like the other ones. We have short, medium and large flipper for each gender.

Let's see all cluster means:
```{r}
cluster_means(cluster$result$centers,cols)
```

##### **Bill_length_mm vs bill_depth_mm**

```{r}
cols=c("bill_length_mm","bill_depth_mm")
cluster <- createCluster(6,25,cols)
cluster$plot
```

We can see six different groups according to length or depth bill.

These are our final centroids:

```{r}
cluster_means(cluster$result$centers,cols)
```

#### 3.2.2 K-medoids

Now, let's try another cluster partitioning model like k-medoids. In this one, each cluster is represented by one of the objects in the cluster. Main difference between k-means and this one is that this takes medoid as the point whose sum of dissimilarities is the lowest. In addition, this algorithm is less sensitive to outliers. Let's define all needed functions:

```{r}
#Function for creating K-Medoids clusters
createKmedoids <- function(k_value,cols){
  set.seed(123) 
  result <- pam(df_scaled[cols], k = k_value, diss = FALSE)
  plot <- fviz_cluster(result, data = df_scaled[cols])
  return (list(result=result,plot=plot))
}
#Function for plotting every algorithm Medoids
plotKmedoidsMeans <- function(k_medoids,cols){
    for (i in 1:nrow(k_medoids)){
    cat("Cluster", i - 1, "->",cols[1],":", k_medoids[i, cols[1]], ",", cols[2],":", k_medoids[i, cols[2]], "\n")
  } 
}

```

We're going to show some of the previous combinations to see if anything is different:

##### **Sex vs body_mass_g**

```{r}
cols <- c("sex","body_mass_g")
cluster <- createKmedoids(6,cols)
cluster$plot
```

As we can see, this algorithm is less sensitive to outliers because in the k-means one, distance to the left upper one was higher.
Let's see final medoids:

```{r}
plotKmedoidsMeans(cluster$result$medoids,cols)
```


##### **Sex vs bill_length_mm**


```{r}
cols <- c("sex","bill_length_mm")
cluster <- createKmedoids(6,cols)
cluster$plot
```

It's similar to K-means but it seems outliers are closer.

Let's see final medoids:

```{r}
plotKmedoidsMeans(cluster$result$medoids,cols)
```

We could classify into small,medium and large for each gender too.


##### **Bill_length_mm vs bill_depth_mm**

Let's finally check if we can group for length and depth for each penguin bill.

```{r}
cols <- c("bill_depth_mm","bill_length_mm")
cluster <- createKmedoids(6,cols)
cluster$plot
```

We can clearly see there are 6 different groups.

Let's see final medoids:
```{r}
plotKmedoidsMeans(cluster$result$medoids,cols)
```


### 3.3 Hierarchical Clustering models
Let's change way to group by, we're going to use hierarchical clustering models now. We're going to use a 50 size random sample:

```{r}
set.seed(123)
indices <- sample(1:dim(df_scaled)[1], 50)
penguin_sample <- df_scaled[indices,]
```


#### 3.3.1 Aglomertive clustering models (hclust)
Let's focus on aglomerative models based in library **hclust**.

##### Complete linkage

```{r}
set.seed(123)
# Complete Linkage
hc_complete <- hclust(dist(penguin_sample), method = "complete")
plot(hc_complete, main = "Complete Linkage")
rect.hclust(hc_complete,k=4)
groups <- cutree(hc_complete,k=4)
print(groups)
```
We can see results in dendogram, with it's bounding boxes and specified groups.

##### Single linkage 
```{r}
set.seed(123)

# Single Linkage
hc_single <- hclust(dist(penguin_sample), method = "single")
plot(hc_single, main = "Single Linkage")
rect.hclust(hc_single,k=3)
groups <- cutree(hc_single,k=3)
print(groups)
```

##### Mean Linkage

```{r}
set.seed(123)

# Mean Linkage (no disponible directamente en hclust, podemos usar 'average' como alternativa)
hc_average <- hclust(dist(penguin_sample), method = "average")
plot(hc_average, main = "Average Linkage")
rect.hclust(hc_average,k=3)
groups <- cutree(hc_average,k=3)
print(groups)
```


#### 3.3.2 Aglomertive clustering model (agnes)
Let's keep developing aglomerative clustering models, but using agnes library now:
```{r}
# Usando agnes para complete linkage como ejemplo
agnes_complete <- agnes(penguin_sample, method = "complete")
plot(agnes_complete, which.plots = 2)

rect.hclust(agnes_complete,k=5)
groups <- cutree(agnes_complete,k=5)
print(groups)
```
#### 3.3.3 Divisive hierarchical clustering (diana)
Let's finally try diana library and it's divisive clustering: 
```{r}
# Clustering divisivo
diana_model <- diana(penguin_sample)
plot(diana_model, which.plots = 2)
rect.hclust(diana_model,k=5)
groups <- cutree(diana_model,k=5)
print(groups)
```


### 4. Conclusions

Our final conclusion is that penguins can be grouped in so many different ways, as it is a subjective grouping method. We could grup them according to sex or other categorical columns, only pulling apart males from females, but you can make different groups such as:

  ·Sex && Body Mass
  ·Sex && Bill length/depth
  ·Sex && Flipper length
  
These are combinations that make more sense for us. We could probably find other grouping criterias but, as we said earlier, it is a subjective task.For example, imagine that you have a male penguin with a determined Bill length, he would probably enter the group of medium long bills males. Given two variables, we could classify them as into new clusters. Previous combinations make more sense because numerical values could be splitted in small, medium or large.We could also have fitted all penguin features into K-means but that would have involved a lot of variables. It could also be a choice to use all variables that you want. We can only see it's divisions with 3 variables(3D graphics).

We have achieved our goals by trying different models, specially different k values in K-Means/K-Medoids.We think partitioning clustering works better with this problem as it is more visual and clear, referring to K-means and K-medoids